---
subtitle: "Injury Recovery & Readiness for Surgery"
params:
  athlete: "Frank Reynolds"
bibliography: references.bib
---

# Phase 1: Recovery from Surgery {#phase-1-recovery-from-surgery}

```{r setup-root, include=FALSE}
knitr::opts_knit$set(root.dir = normalizePath(".."))
```

```{r}
source("Global.R")
source("plot_functions.R")
source("Themes.R")
```

Following ACL reconstructive surgery, the body requires a a period of rest and recovery. Phase 1 is all about optimizing this recovery and working towards satisfying a few simple criteria in order to prepare the knee for more intense progressions in the coming phases.

::: {.callout-important title="Most Important Goals — Phase 1"}
-   Reduce pain and swelling
-   Regain range of motion: Get the knee straight!
-   Re-establish normal gait
-   Re-establish firing of the quadriceps muscles
:::

**`r params$athlete`'s** progress

tion.

```{r}
#Create the table
tbl_phase1 <- criteria_phase1 %>%
  transmute(
    `Outcome Measure` = outcome_measure,
    `Op` = operator_pretty,
    Goal              = goal_pretty,   # nice display string from your sheet
    Score             = score,
    meets
  ) %>%
  gt() %>%
  tab_header(title = md("**Phase 1 Criteria**")) %>%
  fmt_number(
    columns = Score,
    decimals = 0
  ) %>%
  #   fmt_number(
  #   columns = Score,
  #   rows = `Outcome Measure` %in% c("Hamstring Strength", "Quad Strength"),
  #   decimals = 0,
  #   pattern = "{x}%"
  # ) %>%
  sub_missing(columns = Score, missing_text = "") %>%
  cols_label(Op = "") %>%
  cols_width(`Outcome Measure` ~ px(360)) %>%
  cols_align(align = "center", columns = c(Goal, Score)) %>%
  cols_align(align = "right",  columns = Op) %>%
  tab_style(
    style = list(cell_fill(color = "#E8F5E9"),
                 cell_text(color = "#1B5E20", weight = "600")),
    locations = cells_body(columns = Score, rows = meets %in% TRUE)
  ) %>%
  tab_style(
    style = list(cell_fill(color = "#FFF9C4"),
                 cell_text(color = "#7A6A00", weight = "600")),
    locations = cells_body(columns = Score, rows = meets %in% FALSE)
  )  %>%
  cols_hide(meets) %>%
  ak_gt_theme3() %>%
  tab_options(
    table_body.hlines.style = "solid",
    table_body.hlines.width = px(1),
    data_row.padding = px(6)
  )

tbl_phase1
```

```{r}

goals_progress_bar <- function(
    met, total,
    label = "Goals met",
    width = "100%",
    accent = "#D71920",   # table accent (red)
    border = "#A7A9AC",   # table border grey
    track_bg = "#FCFCFC", # header/stripe light bg
    fill_from = "#2E7D32",# green gradient start
    fill_to   = "#4CAF50" # green gradient end
) {
  pct <- if (total > 0) round(100 * met / total) else 0
  
  tagList(
    tags$style(HTML(sprintf("
      .ak-pb-wrap   { margin-top:12px; font-family:'Open Sans',system-ui,-apple-system,Segoe UI,Roboto,Arial,sans-serif; }
      .ak-pb-meta   { display:flex; justify-content:space-between; font-size:0.95rem; margin-bottom:6px; color:#222; }
      .ak-pb-meta strong{ color:%s; font-weight:700; } /* accent like table title */
      .ak-pb       { background:%s; border:1px solid %s; border-radius:10px; height:14px; overflow:hidden; }
      .ak-pb__fill { height:100%%; background:linear-gradient(90deg,%s,%s); transition:width .3s ease; }
    ", accent, track_bg, border, fill_from, fill_to))),
    div(class = "ak-pb-wrap",
        div(class = "ak-pb-meta",
            tags$strong(sprintf("%s: %d/%d", label, met, total)),
            span(sprintf("%d%%", pct))
        ),
        div(
          class = "ak-pb",
          role = "progressbar",
          `aria-valuemin` = "0",
          `aria-valuemax` = "100",
          `aria-valuenow` = pct,
          style = sprintf("width:%s;", width),
          div(class = "ak-pb__fill", style = sprintf("width:%d%%;", pct))
        )
    )
  )
}


# By default: denominator = all criteria (matches your table showing blank scores)
met_n_all   <- sum(criteria_phase1$meets %in% TRUE, na.rm = TRUE)
total_n_all <- nrow(criteria_phase1)

goals_progress_bar(met_n_all, total_n_all)

```

------------------------------------------------------------------------

## What does the research say? {#preop-outcomes}

Recent systematic reviews
