---
#subtitle: "Injury Recovery & Readiness for Surgery"
params:
  athlete: "Frank Reynolds"
bibliography: references.bib
---

# Phase 3: Running, Agility, and Landings {#phase-3-running-agility-and-landings}

```{r setup-root, include=FALSE}
knitr::opts_knit$set(root.dir = normalizePath(".."))
```

```{r}
source("Global.R")
source("plot_functions.R")
source("Themes.R")
```

Once all criteria from Phase 1 have been satisfied, the focus of rehab can shift towards building strength and neuromuscular control.

::: {.callout-important title="Most Important Goals — Phase 3"}
-   Excellent hopping performance
-   Complete agility program
-   Full strength & balance restored
:::

**`r params$athlete`'s** progress

tion.

```{r}
#Create the table
tbl_phase3 <- criteria_phase3_abr %>%
  transmute(
    `Outcome Measure` = outcome_measure,
    `Op` = operator_pretty,
    Goal              = goal_pretty,   # nice display string from your sheet
    Score             = score,
    meets
  ) %>%
  gt() %>%
  tab_header(title = md("**Phase 3 Criteria**")) %>%
  fmt_number(
    columns = Score,
    decimals = 0
  ) %>%
  #   fmt_number(
  #   columns = Score,
  #   rows = `Outcome Measure` %in% c("Hamstring Strength", "Quad Strength"),
  #   decimals = 0,
  #   pattern = "{x}%"
  # ) %>%
  sub_missing(columns = Score, missing_text = "") %>%
  cols_label(Op = "") %>%
  cols_width(`Outcome Measure` ~ px(360)) %>%
  cols_align(align = "center", columns = c(Goal, Score)) %>%
  cols_align(align = "right",  columns = Op) %>%
  tab_style(
    style = list(cell_fill(color = "#E8F5E9"),
                 cell_text(color = "#1B5E20", weight = "600")),
    locations = cells_body(columns = Score, rows = meets %in% TRUE)
  ) %>%
  tab_style(
    style = list(cell_fill(color = "#FFF9C4"),
                 cell_text(color = "#7A6A00", weight = "600")),
    locations = cells_body(columns = Score, rows = meets %in% FALSE)
  )  %>%
  cols_hide(meets) %>%
  ak_gt_theme3() %>%
  tab_options(
    table_body.hlines.style = "solid",
    table_body.hlines.width = px(1),
    data_row.padding = px(6)
  )

tbl_phase3
```

```{r}

goals_progress_bar <- function(
    met, total,
    label = "Goals met",
    width = "100%",
    accent = "#D71920",   # table accent (red)
    border = "#A7A9AC",   # table border grey
    track_bg = "#FCFCFC", # header/stripe light bg
    fill_from = "#2E7D32",# green gradient start
    fill_to   = "#4CAF50" # green gradient end
) {
  pct <- if (total > 0) round(100 * met / total) else 0
  
  tagList(
    tags$style(HTML(sprintf("
      .ak-pb-wrap   { margin-top:12px; font-family:'Open Sans',system-ui,-apple-system,Segoe UI,Roboto,Arial,sans-serif; }
      .ak-pb-meta   { display:flex; justify-content:space-between; font-size:0.95rem; margin-bottom:6px; color:#222; }
      .ak-pb-meta strong{ color:%s; font-weight:700; } /* accent like table title */
      .ak-pb       { background:%s; border:1px solid %s; border-radius:10px; height:14px; overflow:hidden; }
      .ak-pb__fill { height:100%%; background:linear-gradient(90deg,%s,%s); transition:width .3s ease; }
    ", accent, track_bg, border, fill_from, fill_to))),
    div(class = "ak-pb-wrap",
        div(class = "ak-pb-meta",
            tags$strong(sprintf("%s: %d/%d", label, met, total)),
            span(sprintf("%d%%", pct))
        ),
        div(
          class = "ak-pb",
          role = "progressbar",
          `aria-valuemin` = "0",
          `aria-valuemax` = "100",
          `aria-valuenow` = pct,
          style = sprintf("width:%s;", width),
          div(class = "ak-pb__fill", style = sprintf("width:%d%%;", pct))
        )
    )
  )
}


# By default: denominator = all criteria (matches your table showing blank scores)
met_n_all   <- sum(criteria_phase3_abr$meets %in% TRUE, na.rm = TRUE)
total_n_all <- nrow(criteria_phase3_abr)

goals_progress_bar(met_n_all, total_n_all)

```

------------------------------------------------------------------------

## What does the research say? {#preop-outcomes}

Recent systematic reviews

------------------------------------------------------------------------

## Testing

Each card below contains a detailed description of the tests to be conducted during this phase of rehab.

```{r, echo=FALSE, results='asis'}


tests <- criteria_phase3 |>
  dplyr::select(outcome_measure, display_description, additional_information, operator_pretty, goal_pretty, repetitions, calculation) %>%
  mutate(op_goal = paste(operator_pretty, goal_pretty))

# Helper: print a label/value line only if value is non-missing & non-empty
emit_field <- function(label, value) {
  if (length(value) == 0 || all(is.na(value))) return(invisible())
  val <- value[[1]]
  if (is.na(val)) return(invisible())
  val_chr <- trimws(as.character(val))
  if (!nzchar(val_chr)) return(invisible())
  cat("**", label, ":** ", val_chr, "\n\n", sep = "")
}

for (i in seq_len(nrow(tests))) {
  cat('::: {.card .ak-card .has-stripe .accent-primary}\n')       # universal card class
  cat('### ', tests$outcome_measure[i], '\n\n', sep = '')         # heading
  if (!is.na(tests$display_description[i]) &&
      nzchar(trimws(tests$display_description[i]))) {
    cat(tests$display_description[i], '\n\n')
  }
  cat("*", tests$additional_information[i], '\n\n')
  emit_field("Criteria", tests$op_goal[i])                  # always optional-safe
  emit_field("Repetitions", tests$repetitions[i])                # only prints if present
  emit_field("Calculation", tests$calculation[i])

  cat(':::\n\n')
}

```

------------------------------------------------------------------------

## The Program

Below you will find links to your training programs.

[Phase 3: Program 1](https://www.teambuildr.com/)

When considering changes or further individualization of the program, see the lists below highlighting typically safe and unsafe exercises during this phase. To see how these exercises compare to other phases of the rehab, please see the **Exercise Phase Chart** in the book resources. All exercise programming should be done by a CSI Pacific S&C Coach or a physiotherapist.

Exercises and activities during this phase typically include

Peer-reviewed literature that was consulted in the making of pre-op programs is referenced here.

------------------------------------------------------------------------

## Additional Resources

### Blood Flow Restriction

Although, blood flow restrictio

Another more recent study

------------------------------------------------------------------------

## Notes and Considerations for Reviewers

All feedback is welcome, however here are some considerations that should help inform the development of this tool.

-   Please
