---
#subtitle: "Injury Recovery & Readiness for Surgery"
params:
  athlete: "Frank Reynolds"
bibliography: references.bib
---

# Phase 2: Strength & Neuromuscular Control {#phase-1-recovery-from-surgery}

```{r setup-root, include=FALSE}
knitr::opts_knit$set(root.dir = normalizePath(".."))
```

```{r}
source("Global.R")
source("plot_functions.R")
source("Themes.R")
```

Once all criteria from Phase 1 have been satisfied, the focus of rehab can shift towards building strength and neuromuscular control.

::: {.callout-important title="Most Important Goals — Phase 1"}
-   One
-   Two
-   Three
-   Four
:::

**`r params$athlete`'s** progress

tion.

```{r}
#Create the table
tbl_phase2 <- criteria_phase2 %>%
  transmute(
    `Outcome Measure` = outcome_measure,
    `Op` = operator_pretty,
    Goal              = goal_pretty,   # nice display string from your sheet
    Score             = score,
    meets
  ) %>%
  gt() %>%
  tab_header(title = md("**Phase 2 Criteria**")) %>%
  fmt_number(
    columns = Score,
    decimals = 0
  ) %>%
  sub_missing(columns = Score, missing_text = "") %>%
  cols_label(Op = "") %>%
  cols_width(`Outcome Measure` ~ px(360)) %>%
  cols_align(align = "center", columns = c(Goal, Score)) %>%
  cols_align(align = "right",  columns = Op) %>%
  tab_style(
    style = list(cell_fill(color = "#E8F5E9"),
                 cell_text(color = "#1B5E20", weight = "600")),
    locations = cells_body(columns = Score, rows = meets %in% TRUE)
  ) %>%
  tab_style(
    style = list(cell_fill(color = "#FFF9C4"),
                 cell_text(color = "#7A6A00", weight = "600")),
    locations = cells_body(columns = Score, rows = meets %in% FALSE)
  )  %>%
  cols_hide(meets) %>%
  ak_gt_theme3() %>%
  tab_options(
    table_body.hlines.style = "solid",
    table_body.hlines.width = px(1),
    data_row.padding = px(6)
  )

tbl_phase2
```

```{r}

# By default: denominator = all criteria (matches your table showing blank scores)
met_n_all   <- sum(criteria_phase2$meets %in% TRUE, na.rm = TRUE)
total_n_all <- nrow(criteria_phase2)

goals_progress_bar(met_n_all, total_n_all)

```

------------------------------------------------------------------------

## What does the research say? {#sec-p2-research}

### OKC

The training program may include both low/no-load full ROM leg extension as well as ROM-restricted leg extension when the patient is able to lift heavier loads. Minimal ACL strain and reduced PFJ reaction forces have been observed during quadriceps contractions performed between 45-90º of knee flexion, and therefore this is the recommended ROM for loading to be introduced @escamilla2012.

Early introduction of isometrics is recommended ?

Leg Ext isometrics between 90-60º (5x45s, 1-2x per day)

Full ROM (90-0º) against gravity/low loads can be performed once able to comfortably achieve 90º knee flexion

Restricted ROM loaded knee extensions may begin 4 weeks post ACLR (Tempo 3:03, 15-25RM loads)

------------------------------------------------------------------------

## Testing

Each card below contains a detailed description of the tests to be conducted during this phase of rehab.

```{r, echo=FALSE, results='asis'}


tests <- criteria_phase2 |>
  dplyr::select(outcome_measure, display_description, additional_information, operator_pretty, goal_pretty, repetitions, calculation) %>%
  mutate(op_goal = paste(operator_pretty, goal_pretty))

# Helper: print a label/value line only if value is non-missing & non-empty
emit_field <- function(label, value) {
  if (length(value) == 0 || all(is.na(value))) return(invisible())
  val <- value[[1]]
  if (is.na(val)) return(invisible())
  val_chr <- trimws(as.character(val))
  if (!nzchar(val_chr)) return(invisible())
  cat("**", label, ":** ", val_chr, "\n\n", sep = "")
}

for (i in seq_len(nrow(tests))) {
  cat('::: {.card .ak-card .has-stripe .accent-primary}\n')       # universal card class
  cat('### ', tests$outcome_measure[i], '\n\n', sep = '')         # heading
  if (!is.na(tests$display_description[i]) &&
      nzchar(trimws(tests$display_description[i]))) {
    cat(tests$display_description[i], '\n\n')
  }
  cat("*", tests$additional_information[i], '\n\n')
  emit_field("Criteria", tests$op_goal[i])                  # always optional-safe
  emit_field("Repetitions", tests$repetitions[i])                # only prints if present
  emit_field("Calculation", tests$calculation[i])

  cat(':::\n\n')
}

```

------------------------------------------------------------------------

## The Program

Below you will find links to your training programs.

[Phase 2: Program 1](https://www.teambuildr.com/)

When considering changes or further individualization of the program, see the lists below highlighting typically safe and unsafe exercises during this phase. To see how these exercises compare to other phases of the rehab, please see the **Exercise Phase Chart** in the book resources. All exercise programming should be done by a CSI Pacific S&C Coach or a physiotherapist.

Exercises and activities during this phase typically include

Peer-reviewed literature that was consulted in the making of pre-op programs is referenced here.

------------------------------------------------------------------------

## Additional Resources

### Blood Flow Restriction

Although, blood flow restrictio

Another more recent study did not report how early BFR began after surgery, but did use the following criteria to begin BFR: able to stand on one leg unsupported for \>5s, knee ROM between 0-90º, able to perform SLR with sufficient quadriceps activation @barzyk2026 . Patients performed the traditional 30-15-15-15 reps of leg press

------------------------------------------------------------------------

## Notes and Considerations for Reviewers

All feedback is welcome, however here are some considerations that should help inform the development of this tool.

-   Please
